---
title: "XAI3"
author: "Victor Ferrando, Marcos Valero, Marcos Ranchal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(dplyr)
library(randomForest)
library(pdp)
library(ggplot2)
datos<- read.csv("day.csv")
# 1. One-hot encoding para season (3 columnas, eliminamos una para evitar multicolinealidad)
datos <- datos %>%
  mutate(
    season_2 = ifelse(season == 2, 1, 0),
    season_3 = ifelse(season == 3, 1, 0),
    season_4 = ifelse(season == 4, 1, 0)
  )

# 2. Variables MISTY y RAIN
datos <- datos %>%
  mutate(
    MISTY = ifelse(weathersit == 2, 1, 0),
    RAIN = ifelse(weathersit %in% c(3, 4), 1, 0)
  )

# 3. Desnormalizar temp, hum y windspeed
# Según documentación, los valores fueron normalizados dividiendo por:
# temp (C): /41, hum: /100, windspeed: /67
datos <- datos %>%
  mutate(
    temp_real = temp * 41,
    hum_real = hum * 100,
    windspeed_real = windspeed * 67
  )

# 4. Crear days_since_2011
datos$dteday <- as.Date(datos$dteday)
datos$days_since_2011 <- as.numeric(difftime(datos$dteday, as.Date("2011-01-01"), units = "days"))

```

# 1. One dimensional Partial Dependence Plot

```{r rf}
# Select relevant features and target
features <- c("days_since_2011", "temp_real", "hum_real", "windspeed_real")
df_model <- datos[, c(features, "cnt")]

# Fit Random Forest
set.seed(42)
rf_model <- randomForest(cnt ~ ., data = df_model, ntree = 100)#df_model
library(gridExtra)

p1 <- partial(rf_model, pred.var = "days_since_2011", plot = TRUE, rug = TRUE, main = "Effect of Days Since 2011")
p2 <- partial(rf_model, pred.var = "temp_real", plot = TRUE, rug = TRUE, main = "Effect of Temperature")
p3 <- partial(rf_model, pred.var = "hum_real", plot = TRUE, rug = TRUE, main = "Effect of Humidity")
p4 <- partial(rf_model, pred.var = "windspeed_real", plot = TRUE, rug = TRUE, main = "Effect of Wind Speed")

p1
p2
p3
p4
```

The Partial Dependence Plot for "days_since_2011" reveals a generally increasing trend in the predicted number of bike rentals over time. This suggests that, as the days progress from the beginning of 2011, the expected count of rentals grows, likely reflecting factors such as increased adoption of the bike-sharing system, expansion of the service, or seasonal usage patterns. While the trend is not strictly linear—there are periods of slower growth and some plateaus—the overall direction indicates a positive association between time and rental frequency.

The plot for "temp_real" (temperature) shows a clear nonlinear relationship with bike rentals. As temperature increases from low values up to around 25–30°C, the predicted number of rentals rises significantly, indicating that users are more likely to rent bikes in mild to warm weather. However, beyond this optimal temperature range, the rental count begins to decline slightly, suggesting that very hot conditions may deter users due to discomfort or health risks. This aligns well with common behavioral patterns related to outdoor physical activity.

The PDP for "hum_real" (humidity) displays a strong negative effect on bike rentals. While rentals are relatively stable at low to moderate humidity levels, they begin to drop off notably when humidity exceeds 60%, and the decline becomes steeper as humidity continues to rise. This implies that high humidity—often associated with discomfort, sweatiness, or potential rain—reduces the attractiveness of cycling, thereby decreasing the number of rentals predicted by the model.

Finally, the plot for "windspeed_real" (wind speed) also reveals a negative correlation with the number of bike rentals. When wind speed is low (under 10 km/h), the expected number of rentals is high, but as wind speed increases, the predicted rentals steadily decrease. This is likely due to the physical difficulty and reduced safety associated with cycling in windy conditions. At higher wind speeds (above 20 km/h), the number of predicted rentals is significantly lower, confirming that wind acts as a strong deterrent to bike usage




# 2. - Bidimensional Partial Dependency Plot


```{r}
df_model_sampled <- df_model %>% sample_n(300)
set.seed(42)
rf_model_2 <- randomForest(cnt ~ ., data = df_model_sampled, ntree = 100)


# Compute 2D PDP
pdp_2d <- partial(
  object = rf_model_2,
  pred.var = c("temp_real", "hum_real"),
  grid.resolution = 30,
  progress = "none"
)

pdp_df <- as.data.frame(pdp_2d)



ggplot() +
  # Heatmap plano
  geom_tile(
    data = pdp_df,
    aes(x = temp_real, y = hum_real, fill = yhat),
    width = 1.5, height = 2
  ) +
  
  # Escala de color azul
  scale_fill_gradient(low = "#000020", high = "#00bfff") +

  # Rug plots (opcional)
  geom_rug(data = df_model_sampled, aes(x = temp_real), sides = "b", alpha = 0.5) +
  geom_rug(data = df_model_sampled, aes(y = hum_real), sides = "l", alpha = 0.5) +

  # Títulos
  labs(
    title = "2D Partial Dependence Plot",
    x = "Temperature (°C)",
    y = "Humidity (%)",
    fill = "ŷ"
  ) +
  
  # Escalado
  coord_fixed(ratio = 0.25) +
  theme_minimal()



```

The 2D Partial Dependence Plot (PDP) displayed above illustrates the combined effect of temperature (°C) and humidity (%) on the predicted number of bike rentals (ŷ), as estimated by a machine learning model. This type of plot is useful for visualizing how a model's predictions respond to changes in two input variables, while marginalizing over the influence of all other variables in the dataset.

As observed, temperature plays a prominent role in shaping the prediction output. The horizontal color gradient, shifting from dark shades (representing lower predicted rentals) to lighter shades (indicating higher predicted rentals) as temperature increases, clearly shows that higher temperatures are associated with increased bike rental predictions. This trend is consistent with real-world behavior, where people are generally more likely to rent bikes in warmer weather.

In contrast, humidity appears to exert a relatively minor effect on the model’s predictions. The vertical axis, representing humidity, shows minimal color variation from top to bottom, suggesting that changes in humidity alone do not significantly alter the predicted rental count. This may be because humidity either has a limited effect on bike rental decisions or its influence is not strongly captured by the model, potentially due to a non-linear or weak relationship.

The interaction between temperature and humidity also seems limited. While the plot shows distinct horizontal gradients (driven by temperature), there are no corresponding sharp vertical gradients (which would indicate a strong effect from humidity). This indicates that temperature is the dominant variable, and that the combined or interaction effect between temperature and humidity is either weak or negligible within the model.

The rug plots—black tick marks along both axes—represent the distribution of actual data used to train the model. These ticks indicate that the training data is most densely concentrated around 20–30°C and 50–80% humidity. This suggests that the model’s predictions are most reliable within this range, and that caution should be used when interpreting predictions in areas with fewer data points (e.g., temperatures below 10°C or humidity above 90%), as the model may be extrapolating in those regions.

In conclusion, the Random Forest model used for predicting bike rentals has effectively captured a strong and intuitive relationship with temperature, showing that bike rentals increase as temperatures rise. In contrast, humidity has a much weaker or nearly neutral influence, and its interaction with temperature appears minimal. These insights can help inform decision-making related to demand forecasting and planning in bike-sharing systems under different weather conditions.

